---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv

```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv")
```


### Task 3: Calculate the percentage change in each category between Fall 2024 and Fall 2021 using `mutate()`

```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2024` - `Fall 2021`)/`Fall 2021`*100) 
```


### Task 4: Describe our results

Write a few sentences about what you found in the data. What's the story here?

If I were to write a story about UMD Fees I would want to focus on the fact that the total fees have increased more than 18% the last four years. I think it would be cool to do a story where you compare that to a survey of current seniors, who have been here since the beginning of the increase and ask if they think the quality of the services associated with the fees have improved during their time at UMD. That way you can see if the quantitative and qualtitative data is aligned or if the University is making people pay for more than they're receiving.



When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
