---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

What stands out to me about the data is that the salaries in total are relatively low. People think of the White House as a presitigous and highly secure place to work and I think the average person would guess that the compensation for employees would be higher. Particularly when you factor in the cost of living for anyone in the DC area, a salary under $100,000 annually is not very much. The volunteer employees who work for free could cause problems for analysis and skew the data.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize ( mean_salary = mean (salary) )
wh_salaries |> summarize (median_salary = median (salary))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize (mean_salary = mean(salary), median_salary = median(salary), min_salary = min(salary), max_salary = max(salary) )
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean and median tell me that the average salary was a bit higher than I was giving it credit for in my initial analysis of just looking over the data. Both the mean and median are six figure salaries, but still I would argue somewhat on the low side for the lucrative cost of living in the DC area. Like I initially mentioned the people who make zero dollars are outliers which  brings the mean of the data over $3,000 lower than the median. The max salary is also an outlier being the only salary that is over $250,000. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

The best story idea I have would be to talk to some White House employees that are on the lower half of the salary spectrum, about how they afford living in DC on their salary. If any of them have difficult commutes from somewhere like Frederick I think that could be an interesting and novel concept to readers. The next step I'd take would be to look into people that are making from 70-100k and reach out to people who have easily accessible contact information with requests for interviews.
