---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities_crime<-read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities_crime |> mutate (violent_rate_2019 = violent2019/pop2019 * 10000)
md_cities_rates <- md_cities_rates |> mutate (property_rate_2019 = property2019/pop2019 * 10000)
md_cities_rates <- md_cities_rates |> mutate (violent_rate_2020 = violent2020/pop2020 * 10000)
md_cities_rates <- md_cities_rates |> mutate (property_rate_2020 = property2020/pop2020 * 10000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

In 2019 the highest rate of violent crime was in Baltimore City, which is what I would expect, but when you jump to 2020, it becomes Ocean City. I think the general expectation, and what national data showed, was that crime rates went down overall due to lack of population being allowed to go out in public. I think it would be interesting to due a story on the places in Maryland where it went up, particularly with a focus on Ocean City where the rates really went up making it the highest in the whole state. 


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> mutate(property_change = (property_rate_2020-property_rate_2019)/property_rate_2019 * 100)
md_cities_change2 <- md_cities_change |> mutate (violent_change = (violent_rate_2020-violent_rate_2019)/violent_rate_2019 * 100 )
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

I think the patterns are more interesting than the outliers, because some places went from two violent crimes to zero or zero violent crimes to two, which creates a massive percentage change in the rate, but doesn't actually represent a massive difference in the safety of the area. I think the story I would go with is the difference between the change in property rates and violent rates, because it seems like property crime rates decreased in more cities than violent crime rates, which seemed to increase in more places than they didn't. 

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the best story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important. 

Across the nation, lockdowns have been put in place to protect the health of American citizens from the novel coronavirus pandemic, as a result crime rates have dropped in many areas, but not in Ocean City, Maryland. 

The Maryland beach town saw over an 82% increase in violent crime and an almost 17% increase in property crime as compared to 2019 rates.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

##Task 10: Percentiles

```{r}
md_cities_percentiles <- md_cities_change2 |> mutate(violent_percentile = ntile(violent_rate_2020, 100))
md_cities_percentiles |> summarize (iqr_violent_2020 = IQR(violent_rate_2020))
```
I think this data would tweak my previous story idea to widen the scope a little bit. When you look at the counties that are 90th percentile or higher in violent crime rate they are all in rural areas of the state, near the eastern shore or Western Maryland. I think the story could shift to being about how rural counties have the highest rates of crime, despite PG County and Baltimore being considered the most dangerous.
