---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
song_dataset <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
song_dataset |> summarize(mean_danceability=mean(danceability))
song_dataset |> summarize(sd_danceability=sd(danceability))
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
song_dataset |> 
  ggplot() +
  geom_histogram(aes(x=danceability), binwidth=.1 ) +
    geom_vline(aes(xintercept = mean(danceability)), color = "red",linetype = "dashed", size = 1) +
   geom_vline(aes(xintercept = mean(danceability) - sd(danceability)), color = "blue",linetype = "dashed", size = 1) +
geom_vline(aes(xintercept = mean(danceability) + sd(danceability)), color = "blue",linetype = "dashed", size = 1)
```

### Task 5: Add a line to the histogram for the mean danceability score

Take the same code and add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
```


### Task 6: Add two lines to the histogram to show one standard deviation

Take the previous code and add two lines to the histogram for the 1st standard deviation below and above the mean score.

```{r}
```

### Task 7: Describe our results and pitch a story

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and one standard distribution tell us about the songs in the dataset? Draft a short story pitch from this data.

This dataset is distributed somehwat normally, particularly within a standard deviation of the mean, but there are more outliers on the lower end of the danceability scale then the higher. The mean tells us that songs are reaching an average score that is above a 5/10, which would be a logical guess at the average. The standard deviation being relatively normal tells you that there isn't an abnormal amount of variance in the data around the mean.
The best story idea I would have from this would be about current pop music being easy to dance to particularly rap, which I noticed makes up most of the highest scores in the data set. I would like to interview some kind of music critic to see if the data is consistent with the trends he is seeing in music.

### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
