---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
m_and_ms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.

```{r}
sample_5 <- m_and_ms |> sample_n(5)
sample_10 <- m_and_ms |> sample_n(10)
```


### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.

```{r}
sample_10 |> summarize(mean_red=mean(red))
sample_5 |> summarize(mean_red=mean(red))
sample_10 |> summarize(sd_red=sd(red))
sample_5 |> summarize(sd_red=sd(red))
```



### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4

```{r}
m_and_ms |> summarize (mean_red=mean(red))
m_and_ms |> summarize (sd_red=sd(red))
```



### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?

Yes they are similar particularly the full population statistics and the sample of 10 statistics. The sample size of five is further off, but still pretty similar in the grand scheme of things.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

It tells you that you just need a decent size sample to get numbers that are pretty representative of the  real data. The sample of 10 was only 10% of the full data, but it still produced mean and standard deviation numbers that were close enough to be representative of the full population. I think this is easier when your sample is something as binary as red and black M & Ms, but it proves your sample size doesn't need to be that large it just needs to be representative.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
