---
title: "LinearRegression"
output: html_document
---

---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Max Schaeffer"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

We'll do what we did in class, but with a twist: instead of Montgomery County, Pa., you'll be working with data from Montgomery County, Md. Same basic idea, though: linear regression looking at free and reduced-price meals and math test scores.

Just as in the exercise, you will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. You'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores and evaluate the results of your model.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```
### Task: Create a Scatterplot and Describe the Results (5 points)

First, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```
Based on what you see here, what's the relationship between FARMS percentage and math test scores? Compare this scatterplot to the one you made in the class activity.

This is a very different scatterplot then the one from PA that we looked at in class. This shows a much stronger negative relationship betweaen FARMS percentage and test scores, showing the schools with more students getting free lunch are having worse performance on tests.


### Task: Linear Regression Model (7 points)

Next, you'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

Provide the model's output for the three statistics mentioned in the class activity.

P-Value:< 0.00000000000000022 Adjusted R-Squared:0.6734 Coefficient:-0.59511 

Based on what you've learned so far, what does it tell us about the nature of the relationship between FARMS percentage and math scores? Is this model worth exploring further? Why or why not?

These numbers are indicating that there is an at least somewhat statistically significant relationship between the FARMS percentage and the test scores. The P-value tells us we can reject our null hypothesis and our adjusted R-Squared value indicates that the differences in math scores can be largely explained by our change in FARMS percentage, at least in comparison to our much smaller R-Squared value from our in-class activity. These values both give validity to our model and make it worth exploring, 

### Task: Identifying Schools with Biggest Residuals (3 points)

Let's calculate and examine the residuals next. 

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```

Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores? Pick two schools with the largest residuals. Find the zip codes where they are located and using the CensusReporter.org website, describe the most interesting demographic data from each zip code.

I picked Fox Chapel Elementary School which had the largest positive residual, and Germantown Elementary School which had the largest negative residual. Interestingly they're both located in Germantown, MD, so the model is clearly having a difficult time predicting scores in this area. I think the most interesting demographic data of Germantown is the racial and ethnic diversity which is almost exactly even 27% Black, 26% White, 23% Asian, and 19% Hispanic. I wonder if this almost high level of diversity makes the community more difficult to make projections for and if that factored into the models inaccurachy in this case.

## Task: Evaluate the model (5 points)

Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

I would acknowledge that there is statistical validity to the school board members claim, but it is not uniformly accurate and be sure to highlight some of the outliers. I would actually want to take to some of the schools with the positive residuals for my story to see how they're succeeding despite having a high proportion of economically disadvantaged students and see whether this seems replicable for other schools in the county. I would also want to get a local expert voice on why poorer students are having a more difficult time with math in the county.
