---
title: "FinalProject"
output: html_document
---

```{r}
library(tidyverse)
```


```{r}
twentyfive_data<- read_csv("~/Downloads/2025 405 Final Data  - Sheet1.csv")
twentyfour_data <- read_csv("~/Downloads/2024 Final Projectdata - Sheet1.csv")
twentythree_data <- read_csv("~/Downloads/2023 final project data - Sheet1.csv")
```

This is my data separated by year over the last three years. It has basically all stats regarding the ATP's current men's top 50 players serving stats including, but not limited to their percentage of points won on their first and second serve, double fault percentage, and ace percentage. The most important columns for my examination will be percentage of points won on first vs second serve as well as record and win percentage. These will be the backbone of my statistical analysis as the question I'm looking to answer is which serve is it more critical to have success on. I think this data is pretty comprehensive and will give me a good foundation to answer my guiding question. One thing it is missing though is some way to quantify the amount of trophies a player has won, because at the end of the day you can win a lot of matches, but tennis players will always be judged on trophies particularly the grand slams. Luckily it isnt hard to locate that and I can potentially add a column or just factor that data in separately. One question I have: Is there a way to combine the yearly data so it can be used in R, just switching year to year? 

